\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-10-12 document/referee_response.txt

# 
# ------------------- Referee report
# 
# This report presents a review of the article "The X-shooter GRB
# afterglow legacy sample (XS-GRB)" by Selsing and collaborators. The
# authors present and make available to the community a complete sample
# of GRB afterglows observed with the X-Shooter spectrograph on the VLT.
# This dataset is remarkable in several respect (>100 spectra available
# taken over 8 years) and some of these have not be published elsewhere.
# However, there are a number of important issues regarding both the
# scientific measurements and their presentation which are detailed
# below. These concerns need to be addressed before the paper can be
# considered for publication. More details on these points together with
# other comments are provided below.
# 
# # 
# # # 
# Major Concerns:
# 
# - One aspect of these data that needs more emphasis is the intervening
# absorbers. The data presented in this paper present a remarkable
# number of metal and HI intervening absorbers as can be judged from
# reading the appendix. These deserve much more attention, including a
# table listing their redshifts, column density measurements when
# available, metal lines detected, etc. A figure with their redshift
# distribution would also be very welcome. In addition, it would be
# interesting to compare the properties of this population with the
# absorbers towards background quasars. 
#

This paper is meant as a general introduction of the XSGRB legacy sample
and constitutes the first in a series of papers. While we certainly
agree that the issue of intervening absorbers is an interesting one, we
feel that it is best treated in a dedicated publication, which we have
already published Christensen et al., A&A, 608A, 84C, 2017,
(arXiv:1709.01084) and are working on de Ugarte Postigo et al. in
prep.). This paper will contain the complete list of intervening
systems, as well as linelists, redshifts, EWs, etc..

# - The manuscript states that the sample is homogeneous which is a bit
# misleading. The spectrograph used to make the follow-up is the same
# for all afterglows, but the SNR (due to a combination of brightness of
# the source and time of observations after the burst) and the
# rest-frame wavelength coverage vary enormously. While remarkable in
# many respects, it is not appropriate to qualify this sample as
# homogeneous.

The referee is correct here. Our statement about homogeneity concerns
only the selection, not the outcome of our observations. We have
attempted to reword anywhere relevant and we added an explicit caveat on
the resulting S/N homogeneity in paragraph §2.1.

# - There are a great many details provided about the data reduction,
# some of which the reader fails to relate to the final scientific
# goals. For example, a correction is applied to correct offsets in the
# wavelength calibration, which presumably mostly impact the redshift
# determination. How does the improvement translate into the redshift
# determination accuracy? Similarly, for the telluric correction. What
# is the impact of this on the final science? Conversely, a description
# of the flux calibration error estimate would have been welcome.

Concerning the wavelength calibration improvement, we have quantified
the magnitude of the improvement, and explained its relevance (§3.2). We
have also added a short paragraph to §3.6 explaining the impact of the
telluric correction on the inferred results.

Regarding the flux calibration accuracy, we have added a new section
($3.5). This includes a new way to correct for slit losses and tries to
assess the flux accuracy of the spectra presented. The paper has
significantly benefited from this suggestion.

@@

# - The comparison of N(HI) distributions between intervening quasar
# absorbers and intrinsic GRB absorbers is not direct. The measurement
# of the column density of a system in the case of a quasar absorber
# integrate across the whole system. Conversely, in the case of GRBs,
# the burst event takes place at a random position in the host galaxy
# and in fact, on average, the column density will measure over only
# half of the system. This geometric effect should be taken into
# account.

This is certainly true, when comparing QSO DLAs to GRB DLA. We have
updated the figure and text surrounding Fig. 10 to reflect this. For
Fig. 9, where we are comparing GRB DLAs to GRB DLA, we have retained the
original values for N_HI.

# - Another concern is to do with using absorption features to measure
# the redshift. Indeed, one cannot exclude that the absorption is not at
# the position of host galaxy but lying in front of it. For this reason,
# the redshifts provided should be seen as upper limits. 
# 
# 
# 
# 
# # 
# Other Concerns:
# 
# - the links to the actual data seem to be inactive. It is important to
# be able to review the material which is made available to the
# community.

Completely true. These are now active. Especially the ESO phase 3
material have taken some time to get through. Apologies.

# - caption of Fig. 3 is cut-out

This is now fixed.

# - Fig. 5: please label the y-axis

Done, thanks.

# - Fig. 10: caption is cut-out again

Fixed, thanks.

# GRB150403A: could the continuum be lower?

We have remeasured the N_HI for this burst. It does not change the
inferred N_HI value, but the continuum now better matches the
measurement.

# GRB150915A: is this fit significant?

We agree this could be misinterpreted. This bursts has Lyman alpha
emission in the trough, as indicated in Tab. 4. We have now marked all
the bursts affect by Lyman alpha emission.

# - appendix A: refer to the spectral PSF in the title?

This has now been reworded.

# - appendix B: science-verified, Paranalization: perhaps avoid this ESO
# jargon?

We have tried to weed out the use of these types of words.

# typos: - pt. 4 p. 3: Paranl - p. 3: exampl - p. 3: sorrounding - p. 9:
# curios - there are numerous typos in the appendix which I will not
# review here. Please make sure to double check spelling there.

Thanks for noting these. They have been fixed.

# I would be happy to review an updated version of the manuscript where
# the above comments are taken into account.
# 
# -----------------------

### 2018-10-12 document/XSGRB_sample_rev1.tex
Finishing referee response

@@

The \textit{Neil Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of the
intergalactic medium, which was already hinted at by results from earlier
missions \citep[e.g.,][]{Paradijs2000, Ricker2004}. With the three on-board
instruments, the Burst Alert Telescope \citep[BAT;][]{Barthelmy2005}, the X-Ray
Telescope \citep[XRT;][]{Burrows2005}, and the UltraViolet and Optical Telescope
\citep[UVOT;][]{Roming2005}, \swift~is an ideal observatory for GRB hunting. A
crucial aspect of the success of the {\it Swift} mission has been the extensive
ground-based follow-up observations of the afterglows and of the host galaxies
of the GRBs, involving a large community of researchers. This fruitful
collaboration has been facilitated by the open data access policy of the {\it
	Swift} mission. The close collaboration between detection facilities and
electromagnetic  follow-up campaigns continue to be immensely rewarding, as
recently highlighted by the simultaneous detection of gravitational waves and
light from the neutron star merger in the shape of GW170817/GRB170817A/AT2017gfo
\citep{LIGOScientificCollaboration2017a, LIGOScientificCollaboration2017}.

@@

\begin{enumerate}
	\item The GRB triggered \textit{Swift} onboard.
	\item The X-ray position must be available less than 1 hr after the GRB.
	%    \item X-shooter is the currently operating instrument at the telescope.
	% DM I removed the line above because we do send the trigger even if X-shooter is not mounted (though, the trigger is rejected, as explained in the text above)
	%    \item GRB promptly visible from Cerro Paranal.
	%    \item Error radius < 60\arcsec.
	%    \item No limit on declination.
	%    \item No constraint on Moon phase or distance.
	\item The elevation of the source in the sky is $> 22^\circ$ (both at trigger time and 15 min after).
	\item The Sun elevation from Paranl is $< -12^\circ$ (both at trigger time and 15 min after).
	\item Galactic $A_V \leq 1.0$~mag according to the maps of \citet{Schlegel1998}.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}.
\end{enumerate}

@@

The use of RRM is unique as it allows to sample a long logarithmic time span in
the GRB lifetime, and it exploits the extreme brightness of early afterglows. A
pivotal exampl is the study of temporal variability of GRB afterglow absorption
systems due to effect of the GRB itself on the sorrounding medium \citep[e.g.,
see][]{Dessauges-Zavadsky2006, Vreeswijk2007, DElia2009, Vreeswijk2013}.

@@

@@

\begin{figure*}
	\centerline{\includegraphics[width=\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$R$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption and emission lines, including
	absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
	$\sim$ 4000 $\AA$ is due to \lya~in the host. We have marked the most prominent
	lines seen in GRB afterglows from \citet{Christensen2011a}. The regions of most
	severe telluric absorption are highlighted by grey-shading the background.
	Additionally, three intervening systems are seen in this sightline. This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
\label{fig:spectrum}
\end{figure*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Slitloss correction and flux calibration accuracy} \label{fluxes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For the majority of the observation presented in this sample, the slit width of
the target observations is less than the seeing PSF. This means that a fraction
of the incoming light is lost outside the slit. As an example, for atmospheric
conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a
0\farc9 slit, only 76.1 per cent of the light will enter the slit. This slitloss
is not corrected by the flux calibration, as flux standard stars are always
observed with a 5~\arcsec slit. As a consequence, almost all spectra will,
without slitloss correction, underestimate the flux by some fraction.
Additionally, slitlosses are wavelength dependent due to the wavelength
dependence of seeing \citep{Boyd1978} and as the seeing changes across the slit
in wavelength, so does the slitloss. This artificially introduces a change in
continuum slope.

To correct for the slitlosses, we develop an algorithm that can, based on the
atmospheric conditions at time of observation and the wavelength dependence of
seeing, estimate the losses at each wavelength bin. We start with the DIMM
seeing \citep{Sarazin1990}, which measures the seeing FWHM at zenith. We correct
the zenith seeing by the airmass of the observational sightline, which gives us
the effective seeing, delivered outside the slit at 5000 \AA. We can then use
that the seeing scales as $\sim \lambda^{-1/5}$ to obtain a measure of the
seeing across the entire wavelength coverage of X-shooter. At each wavelength
bin of the spectra, we simulate a synthetic 2D seeing PSF based on a normalised,
2D Moffat profile which we integrate over the used slit width. The inverse of
the fraction of the light that enters the slit, can then be used to correct for
the estimated slit losses at each wavelength bin.

To assess the accuracy of the flux-calibration, after the slit loss correction,
we compare a subset of afterglow spectra with photometry derived from the
acquisition camera. The acquisition camera takes a short image before
integration causing the delay between the image and the spectroscopic
integration is negligible. The acquisition camera suffer from varying imaging
quality, lacking calibration files, and regions of reduced thoughput due to
contaminants on the CCD. The derived photometry therefore suffers from
systematics that can be hard to correct for. Because the afterglow brightness
changes rapidly with time, no other imaging are sufficiently close in time to
use.

We select all GRBs for which the aquisition camera used a Sloan filter
\citep{Gunn1998} and is covered by either SDSS \citep{Abolfathi2018} or
Pan-STARRS \citep{Chambers2016}. Additionally we select acquisition camera
magnitude brighter than 21.5 mag, and delay-time between GRB trigger and
spectroscopic observation < 10 hours. These cuts leaves 7 afterglow spectra, for
which a decent calibration of the photometric zero point can be derived. Using
this set we can make a test of the final flux-calibration accuracy.

To compare the acquisition camera photometry with the spectra, we calculate the
synthetic magnitude of the spectrum, in the used passband, along with the
associated error. These two measures should be the same, modulo systematic
differences affecting the spectra - i.e. bag pixels/regions etc.

@@

In this section, we describe the efficiency of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
obtained sample is representative for the full \textit{Swift} sample. An
important note is that here we provide a release for \textit{all} GRBs after
\startdate, that have been observed with X-shooter, while only a subset of these
constitutes our \textit{statistical} sample. The statistical sample is based on
the selection criteria described in Sect. \ref{samplecrit}. Some bursts not
fulfilling the sample criteria have been followed up due to interesting
characteristics, e.g., curios properties of their light curves, their
brightness, etc. These bursts are not discussed as part of the investigation of
the statistical properties of the GRB population. A prime example of a
spectacular burst outside the statistical sample is the bright \textit{INTEGRAL}
burst GRB~161023A (de Ugarte Postigo et al., in prep), that contains at least 15
intervening absorption systems (See \ref{161023}).

@@

@@

In Fig. \ref{fig:NH_dist}, we also show the column density distribution
for the 12081 quasar absorbers with \nh~$> 20$ from \citet{Noterdaeme2012b}. The
fact that GRBs are systematically located behind the highest \nh, previously
noted \citep[e.g.,][]{Prochaska2007, Fynbo2009}, is very clear in this
figure. The reason for this is that quasar sample sight-lines through galaxies
that are cross-section selected, whereas GRB sight-lines probe the dense,
star-forming regions in their hosts.

\begin{figure}[!t]
	\centering \includegraphics[width=\columnwidth]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citet{Noterdaeme2012b} in blue. Overplotted in
	green is the kernel density estimate of absorbers in GRB sightlines. Values are
	taken from the compilation in Tanvir et al. (submitted), along with the new
	values presented in this sample. We also show in the red the values derived in
	this work. The marginal distributions for the three samples are also shown
	along the left side and on the top, where the different environments probed are
	clearly visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}.} \label{fig:NH_dist}
\end{figure}

@@

\begin{figure*}[!h]
	\centering
	\includegraphics[page=1, width=\linewidth]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column densities for all bursts with a
		clear Lyman alpha absorption system. In solid black is shown the spectrum with
		black dotted giving the corresponding 1-$\sigma$ error. The black dashed line shows zero
		flux density. The solid red line is the absorption of column density equal to
		the value presented in Tab. \ref{tab:HI} with the 1-$\sigma$ interval shown
		with dashed lines.}
	\label{fig:HI1}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=2, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=3, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=4, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=5, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=6, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI6}
\end{figure*}
\clearpage

@@

Observed during paranalization of the instrument, these data forms the basis of
GCN 15571 \citep{GCN15571}, but are not published elsewhere. Observations began
20.6 hours after the BAT trigger and 4.8 ks spectroscopic integration in all
three arms reveals the absorption signature for a host at $z = 1.266$ from the
detection of \mgii, \mgi, \SIii, \feii, \aliii. Because this burst was observed
before the instrument was science-verified, it does not enter into the
statistical sample.

@@

Observed during the first science verification period and was reported in GCN
9761 \citep{GCN.9761} and is additionally used as the basis for the master
thesis by \'Asa Sk\'ulad\'ottir (2010). 7.2 ks integration starting 10.2 hours
after the GRB trigger yields a clear afterglow continuum in all arms. The
simultaneous detection of absorption lines identified as \lya, \SIii, \oi,
\SIi$^*$, \SIiv, \civ, \feii, \alii, \aliii~and \mgii~at $z = 2.737$ sets it as
the redshift of the GRB. Because this burst is observed before the instrument
was science-verified, it does not enter into the statistical sample.

@@

Obtained during the second science
verification period, this dataset forms the basis of GCN 9942 \citep{GCN.9942} and is
additionally published in \citet{DElia2010}. Spectroscopic integration started
22 hours after the BAT trigger and from the acquisition camera the optical
afterglow is measured to R = 17.9 mag at the beginning of the observations which
causes a strong continuum to be seen in all arms. An absorption trough due to
\lya~is clearly visible along with numerous metal resonance lines \civ, \SIii,
\SIii$^*$ \feii, \mgii, all at $z = 2.106$. Because this burst was observed
before the instrument was science-verified, it does not enter into the
statistical sample.

@@

The first burst observed during normal operation after science verification was
completed and the first burst that enters the statistical sample. These data are
the basis for GCN 10042 \citep{GCN10042} and is published in
\citet{Wiersema2012}. With a bright afterglow and a rapid follow-up, this
spectrum is of pristine quality. The afterglow continuum is bright throughout
all spectroscopic arms which allows the ready detection of \alii, \aliii, \feii,
\mnii, \mgii, \mgi, and \caii~- all located at $z = 0.971$, setting it as the
redshift of the GRB.

>>>>

\usepackage{caption}

@@

The \textit{Neil Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of the
intergalactic medium, which was already hinted at by results from earlier
missions \citep[e.g.,][]{Paradijs2000, Ricker2004}. With the three on-board
instruments, the Burst Alert Telescope \citep[BAT;][]{Barthelmy2005}, the X-Ray
Telescope \citep[XRT;][]{Burrows2005}, and the UltraViolet and Optical Telescope
\citep[UVOT;][]{Roming2005}, \swift~is an ideal observatory for GRB hunting. A
crucial aspect of the success of the {\it Swift} mission has been the extensive
ground-based follow-up observations of the afterglows and of the host galaxies
of the GRBs, involving a large community of researchers. This fruitful
collaboration has been facilitated by the open data access policy of the {\it
	Swift} mission. The close collaboration between detection facilities and
electromagnetic  follow-up campaigns continue to be immensely rewarding, as
recently highlighted by the simultaneous detection of gravitational waves and
light from the neutron star merger in the shape of GW170817/GRB170817A/AT2017gfo
\citep{2017ApJ...848L..14G, 2017ApJ...848L..15S, 2017ApJ...848L..16S, 2017ApJ...848L..17C, 2017ApJ...848L..18N, 2017ApJ...848L..19C, 2017ApJ...848L..27T, 2017ApJ...850L...1L, 2017Natur.551...64A, 2017Natur.551...67P, 2017Natur.551...71T, 2017Natur.551...75S, 2017Sci...358.1583K, LIGOScientificCollaboration2017a, LIGOScientificCollaboration2017}.

@@

\begin{enumerate}
	\item The GRB triggered \textit{Swift} onboard.
	\item The X-ray position must be available less than 1 hr after the GRB.
	%    \item X-shooter is the currently operating instrument at the telescope.
	% DM I removed the line above because we do send the trigger even if X-shooter is not mounted (though, the trigger is rejected, as explained in the text above)
	%    \item GRB promptly visible from Cerro Paranal.
	%    \item Error radius < 60\arcsec.
	%    \item No limit on declination.
	%    \item No constraint on Moon phase or distance.
	\item The elevation of the source in the sky is $> 22^\circ$ (both at trigger time and 15 min after).
	\item The Sun elevation from Paranal is $< -12^\circ$ (both at trigger time and 15 min after).
	\item Galactic $A_V \leq 1.0$~mag according to the maps of \citet{Schlegel1998}.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}.
\end{enumerate}

@@

The use of RRM is unique as it allows to sample a long logarithmic time span in
the GRB lifetime, and it exploits the extreme brightness of early afterglows. A
pivotal example is the study of temporal variability of GRB afterglow absorption
systems due to effect of the GRB itself on the surrounding medium \citep[e.g.,
see][]{Dessauges-Zavadsky2006, Vreeswijk2007, DElia2009, Vreeswijk2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Slit loss correction and flux calibration accuracy} \label{fluxes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For the majority of the observations presented in this sample, the slit width of
the target observations is less than the seeing PSF. This means that a fraction
of the incoming light is lost outside the slit. As an example, for atmospheric
conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a
0\farc9 slit, only 76.1 per cent of the light will enter the slit. This slit loss
is not corrected by the flux calibration, as flux standard stars are always
observed with a 5\arcsec{} slit. As a consequence, almost all spectra will,
without slit loss correction, underestimate the flux by some fraction.
Additionally, slit losses are wavelength dependent due to the wavelength
dependence of seeing \citep{Boyd1978} and as the seeing changes across the slit
in wavelength, so does the slit loss. This artificially introduces a change in
continuum slope.

\begin{figure}[!t]
	\centerline{\includegraphics[width=\columnwidth]{figures/flux_accuracy}}
\caption{The left panel shows the acquisition camera magnitudes as a function
	of synthetic photometry derived from the spectra. The blue, dashed line
	represent equality between the two measures. The right panel shows a histogram
	of the differences between the two. The vertical lines in the histogram
	indicates the 16th, 50th, and 84th percentiles of the distribution ($\Delta
	\mathrm{mag} = 0.07 ^{+0.51} _{-0.45} $). This shows the accuracy of the flux
	calibration. The colouring of the data-points shows the photometric
	passband used by the acquisition camera, as indicated by the labels. The $R$- and $I$- band magnitudes have been calibrated against the USNO-B1 catalog,
	which has a poor photometric accuracy; this contributes significantly to the  observed scatter, and for this reason we indicate these data with a light color.} \label{fig:flux_acq}
\end{figure}

To correct for the slitlosses, we develop an algorithm that can, based on the
atmospheric conditions at time of observation and the wavelength dependence of
seeing, estimate the losses at each wavelength bin. We start with the DIMM
seeing \citep{Sarazin1990}, which measures the seeing FWHM at zenith. We correct
the zenith seeing for the airmass of the observational sightline, which gives us
the effective seeing, delivered outside the slit at 5000 \AA. We can then use
that the seeing scales as $\sim \lambda^{-1/5}$ to obtain a measure of the
seeing across the entire wavelength coverage of X-shooter. At each wavelength
bin of the spectra, we simulate a synthetic 2D seeing PSF based on a normalised,
2D Moffat profile which we integrate over the used slit width. This can be used to correct for the estimated slit losses at each wavelength bin.

To assess the accuracy of the flux-calibration, after the slit loss correction,
we compare a subset of afterglow spectra with photometry derived from the
acquisition camera. The acquisition camera takes a short image just before
the spectroscopic integration, which only introduces a minimal delay and allows us to secure photometry for the vast majority of events. As photometric flux standards are rarely acquired, we resort to sky surveys such as the USNO catalog (for the $R$ and $I$ bands; \citealt{Monet2003}) or SDSS and Pan-STARRS ($griz$; \citealt{Abolfathi2018,Chambers2016}), where available.

To compare the photometry with the spectral flux calibration,
we calculate synthetic magnitudes from the spectrum, in the used passbands,
along with the associated error. Using this set we can make a test of the final flux-calibration accuracy, which
we show in Fig.~\ref{fig:flux_acq}. As can be seen, there is a general agreement
between the flux-calibration and the derived acquisition magnitudes, although
with significant scatter. The right panel in Fig.~\ref{fig:flux_acq} shows a
histogram of the differences, which can be summarized with $\Delta \mathrm{mag}
= 0.07 ^{+0.51} _{-0.45} $, where the errors are the 16th and 84th percentiles of
the distribution. The scatter is smaller if we exclude measurements calibrated against the USNO catalog, which is known to have a poor accuracy, also because, given the small field-of-view, we often need to rely of faint reference stars.

Similar to what is found in \citet{Lopez2016}, there is in some cases a mismatch
in the flux-calibration between the spectroscopic arms. It is outside the scope
of this work to correct for these offsets on an individual burst basis.The flux
calibration of the spectra presented here shows a general good agreement with
photometry although with some scatter. The spectral shape within the
individual arms is correct, but the absolute scaling might in some cases be
discrepant.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
	\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$
		that illustrates the typical data quality. The continuum estimate is shown in
		dashed red and the error spectrum in solid blue. The acquisition magnitude is
		$R$ = 20, meaning it is in the brighter end of the sample presented here, but
		not the brightest. The spectrum is rich in absorption and emission lines, including
		absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
		$\sim$ 4000 $\AA$ is due to \lya~in the host. We have marked the most prominent
		lines seen in GRB afterglows from \citet{Christensen2011a}. The regions of most
		severe telluric absorption are highlighted by grey-shading the background.
		Additionally, three intervening systems are seen in this sightline. This
		spectrum has been analyzed in detail in \citet{Friis2015}.}
	\label{fig:spectrum}
\end{figure*}

@@

@@

In this section, we describe the efficiency of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
obtained sample is representative for the full \textit{Swift} sample. An
important note is that here we provide a release for \textit{all} GRBs after
\startdate, that have been observed with X-shooter, while only a subset of these
constitutes our \textit{statistical} sample. The statistical sample is based on
the selection criteria described in Sect. \ref{samplecrit}. Some bursts not
fulfilling the sample criteria have been followed up due to interesting
characteristics, e.g., curious properties of their light curves, their
brightness, etc. These bursts are not discussed as part of the investigation of
the statistical properties of the GRB population. A prime example of a
spectacular burst outside the statistical sample is the bright \textit{INTEGRAL}
burst GRB~161023A (de Ugarte Postigo et al., in prep), that contains at least 15
intervening absorption systems (See \ref{161023}).

@@

A possible complication is that a large fraction of the redshifts are 
measured from absorption systems, which in principle only provide lower 
limits to the GRB redshift. While for individual systems we cannot rule 
out a low-column density host which is missed, especially in low S/N 
spectra (see for example the discussion by 
\citealt{2008ApJ...688..470P, Perley2017a}), we argue that this issue is very 
unlikely to affect significantly our sample.

First, numerous works, including this one, have found that GRB 
absorption systems have much higher column densities than randomly 
selected sight lines 
\citep{2001ApJ...549L.209G,Prochaska2007, Fynbo2009}. 
This fact itself shows that the bulk of GRB absorption lines are not due 
to random intervening systems, but are peculiar to the GRB host. One 
second line of evidence comes from the systematic detection of nebular 
emission lines at the same redshift as the absorption system. 
\citet{Kruhler2015} report a 94\% detection rate of emission 
lines at $z < 1$ (where sensitivity is not an issue, even in the 
presence of a bright afterglow), always in agreement with absorption. 
Besides, it is extremely rare to detect emission from QSO intervening 
systems (see e.g. \citealt{2010MNRAS.408.2128F} for a discussion), which 
again indicates that GRB absorption systems are native to their hosts. 
An even more direct determination comes from the detection, in several 
cases, of fine-structure and metastable lines (for example from Fe and 
Ni), which are exclusively detected in GRB environments (and have been 
directly connected to the UV field generated by the GRB itself; e.g. 
\citealt{2006ApJ...648...95P, Vreeswijk2007, DElia2009}). 
Finally, at $z \gtrsim 2$ (where Ly$\alpha$ enters the observed range) 
the onset of the Ly$\alpha$ forest also provides an \textit{upper} limit 
on the GRB redshift, which again is always consistent with the 
absorption line redshift. The whole set of these arguments indicates 
that in only very few cases, if any at all, is a misidentification of 
the redshift with the intervening system a concrete possibility.

@@

In Fig. \ref{fig:NH_dist}, we also show the column density distribution
for the 12081 quasar absorbers with \nh~$> 20$ from \citet{Noterdaeme2012b}. The
fact that GRBs are systematically located behind the highest \nh, previously
noted \citep[e.g.,][]{Prochaska2007, Fynbo2009}, is very clear in this
figure. The reason for this is that quasar sample sight-lines through galaxies
that are cross-section selected, whereas GRB sight-lines probe the dense,
star-forming regions in their hosts. When comparing the GRB DLAs and and QSO DLAs we correct for the fact that, on average, the GRB light only goes though half the DLA. In Fig. \ref{fig:NH_dist}, we have therefore multiplied the inferred $N_{\mathrm{HI}}$ values by 2 \citep{Prochaska2008}.

\begin{figure}[!t]
	\centering \includegraphics[width=\columnwidth]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citet{Noterdaeme2012b} in blue. Overplotted in
	green is the kernel density estimate of absorbers in GRB sightlines. Values are
	taken from the compilation in Tanvir et al. (submitted), along with the new
	values presented in this sample. We also show in the red the values derived in
	this work. The marginal distributions for the three samples are also shown
	along the left side and on the top, where the different environments probed are
	clearly visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}. To correct for geometric differences between GRB DLAs and QSO DLAs, we have multiplied the inferred GRB DLA $N_{\mathrm{HI}}$ by 2.} \label{fig:NH_dist}
\end{figure}

@@

\begin{figure*}[!h]
	\centering
	\includegraphics[page=1, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column densities for all bursts with a
		clear Lyman alpha absorption system. In solid black is shown the spectrum with
		black dotted giving the corresponding 1-$\sigma$ error. The black dashed line shows zero
		flux density. The solid red line is the absorption of column density equal to
		the value presented in Tab. \ref{tab:HI} with the 1-$\sigma$ interval shown
		with dashed lines.}
	\label{fig:HI1}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=2, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=3, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=4, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=5, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=6, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI6}
\end{figure*}
\clearpage

@@

Observed during the verification of the instrument, these data forms the basis of
GCN 15571 \citep{GCN15571}, but are not published elsewhere. Observations began
20.6 hours after the BAT trigger and 4.8 ks spectroscopic integration in all
three arms reveals the absorption signature for a host at $z = 1.266$ from the
detection of \mgii, \mgi, \SIii, \feii, \aliii. Because this burst was observed
before the instrument was finished completely, it does not enter into the
statistical sample.

@@

Observed during the first instrument testing period and was reported in GCN
9761 \citep{GCN.9761} and is additionally used as the basis for the master
thesis by \'Asa Sk\'ulad\'ottir (2010). 7.2 ks integration starting 10.2 hours
after the GRB trigger yields a clear afterglow continuum in all arms. The
simultaneous detection of absorption lines identified as \lya, \SIii, \oi,
\SIi$^*$, \SIiv, \civ, \feii, \alii, \aliii~and \mgii~at $z = 2.737$ sets it as
the redshift of the GRB. Because this burst is observed before the instrument
was science-verified, it does not enter into the statistical sample.

@@

Obtained during the second instrument testing period, this dataset forms the
basis of GCN 9942 \citep{GCN.9942} and is additionally published in
\citet{DElia2010}. Spectroscopic integration started 22 hours after the BAT
trigger and from the acquisition camera the optical afterglow is measured to R =
17.9 mag at the beginning of the observations which causes a strong continuum to
be seen in all arms. An absorption trough due to \lya~is clearly visible along
with numerous metal resonance lines \civ, \SIii, \SIii$^*$ \feii, \mgii, all at
$z = 2.106$. Because this burst was observed before the instrument was
science-verified, it does not enter into the statistical sample.

@@

The first burst observed during normal operation and the first burst that enters
the statistical sample. These data are the basis for GCN 10042 \citep{GCN10042}
and is published in \citet{Wiersema2012}. With a bright afterglow and a rapid
follow-up, this spectrum is of pristine quality. The afterglow continuum is
bright throughout all spectroscopic arms which allows the ready detection of
\alii, \aliii, \feii, \mnii, \mgii, \mgi, and \caii~- all located at $z =
0.971$, setting it as the redshift of the GRB.
